**A framework for effective application of machine learning to a microbiome-based classification problem**

### 2020-01-27 submission/manuscript.Rmd
tiff figures

---
title: '**Effective application of machine learning to a microbiome-based classification problem**'
csl: mbio.csl
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
bibliography: references.bib
---

@@

Running title: Machine learning methods applied to a microbiome study

@@

As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose disease. However, the structure of the human microbiome is remarkably variable among individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models. This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals. Furthermore, a growing number of studies have shown that it is rare for a single bacterial species to be associated with a disease. Instead, subsets of the microbiome account for differences in health. Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population separately [@segata_metagenomic_2011]. Recently, machine learning (ML) models have grown in popularity among microbiome researchers as our ability to sample large numbers of individuals has grown; such models can effectively account for the interpersonal microbiome variation and the ecology of disease.

@@

In this study, we provided steps toward standardization of machine learning methods for microbiome studies which are often poorly documented and executed. To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing the predictive performance, interpretability and training times of seven modeling approaches with the same dataset and pipeline. We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel. We also trained four non-linear models: SVM with radial basis function kernel, a decision tree, random forest, and gradient boosted trees. We compared their predictive performance, interpretability, and training time. To demonstrate the performance of these modeling approaches and our pipeline on a case study, we used a data from a previously published study that sought to classify individuals as having healthy colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016]. This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. With this dataset, we developed an ML pipeline that can be used in many different scenarios for training and evaluating models. This framework can be easily applied to other host-associated and environmental microbiome datasets. We also provided an aspirational rubric for evaluating the rigor of ML practices applied to microbiome data [Figure S1] to urge the audience to be diligent with with their study design, model selection, development, evaluation and interpretation.

@@

Model selection requires tuning hyperparameters. Hyperparameters are parameters that need to be specified or tuned by the user, in order to train a model for a specific modeling problem. For example, when using regularization, C is a hyperparameter that indicates the penalty for overfitting. Hyperparameters are tuned using the training data to find the best model. We selected hyperparameters by performing repeated five-fold cross-validation (CV) on the training set [Figure 1]. The five-fold CV was also stratified to maintain the overall case and control distribution. We chose the hyperparameter values that led to the best average CV predictive performance using the area under the receiver operating characteristic curve (AUROC) [Figure S1 and S2]. The AUROC ranges from 0, where the model’s predictions are perfectly incorrect, to 1.0, where the model perfectly distinguishes between cases and controls. An AUROC value of 0.5 indicates that the model's predictions are no different than random. To select hyperparameters, we performed a grid search for hyperparameter settings when training the models. Default hyperparameter settings in developed ML packages available in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and need to be optimized for each new ML task. For example, L1-regularized SVM with linear kernel showed large variability between different regularization strengths (C) and benefited from tuning as the default C parameter was 1 [Figure S1].

@@

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having healthy colons or SRNs [Figure 2]. The predictive performance of random forest model was higher than other ML models with a median  `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`], though not significantly (p=0.5; The p-value was manually calculated using the sampling distribution of the test statistic under the null hypothesis) (Figure S3). Similarly, L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(median_iqr(all, "L2_Logistic_Regression")[1])` [IQR `r paste(median_iqr(all, "L2_Logistic_Regression")[2])`-`r paste(median_iqr(all, "L2_Logistic_Regression")[3])`], `r paste(median_iqr(all, "XGBoost")[1])` [IQR `r paste(median_iqr(all, "XGBoost")[2])`-`r paste(median_iqr(all, "XGBoost")[3])`], `r paste(median_iqr(all, "L2_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L2_Linear_SVM")[2])`-`r paste(median_iqr(all, "L2_Linear_SVM")[3])`] and `r paste(median_iqr(all, "RBF_SVM")[1])` [IQR `r paste(median_iqr(all, "RBF_SVM")[2])`-`r paste(median_iqr(all, "RBF_SVM")[3])`], respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(median_iqr(all, "L1_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L1_Linear_SVM")[2])`-`r paste(median_iqr(all, "L1_Linear_SVM")[3])`] and `r paste(median_iqr(all, "Decision_Tree")[1])` [IQR `r paste(median_iqr(all, "Decision_Tree")[2])`-`r paste(median_iqr(all, "Decision_Tree")[3])`], respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and L2-regularized SVM with linear kernel) performed nearly as well as non-linear models.

@@

To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S4]. The variation in cross-validation performance within both models at smaller sample sizes was larger than when the full collection of samples was used to train and validate the models. Because of the high dimensionality of the microbiome data (6920 OTUs), large sample sizes can lead to better models.

__Interpretation of each ML model.__  To evaluate the overall structure of the CRC dataset and show that the structure of the human microbiome is remarkably variable among individuals with SRNs, we analyzed the Bray-Curtis distances with an NMDS ordination [Figure S5]. The lack of seperation between healthy patients and patients with SRNs, show that traditional statistical approaches do not help us to adequately address disease ecology. This is why we often use ML models not just to predict a health outcome, but also to identify potential biomarkers for disease. Therefore, model interpretation becomes crucial for microbiome studies. Interpretability is related to the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017; @ribeiro_why_2016; @nori_interpretml:_2019]. ML models often decrease in interpretability as they increase in complexity. In this study, we used two methods to help interpret our models.

@@

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S7]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S6], 17 of the 20 top OTUs (e.g., OTU 609, 822, 1239) were deemed important by both interpretation methods. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods. These results indicate that both methods are consistent in selecting the most important OTUs.

@@

**Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1). We compared the performance of the models pairwise by calculating the difference between AUROC values from the same data-split (for 100 data-splits). We determined if the models were significantly different by calculating the empirical p-value  (2 x min(% of AUROC differences $\geq$ 0, % of AUROC differences $\leq$ 0) for the double tail event (e.g., Figure S3).

@@

\newpage
\includegraphics{Figure_1.pdf}
**Figure 1. Machine learning pipeline.**   We split the data to create a training (80%) and held-out test set (20%). The splits were stratified to maintain the overall class distribution. We performed five-fold cross-validation on the training data to select the best hyperparameter setting and then used these hyperparameters to train the models. The model was evaluated on the held-out data set. Abbreviations: cvAUC, cross-validation area under the receiver operating characteristic curve.
\newpage
\includegraphics{Figure_2.png}

**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross-validation and testing performances.** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by a horizontal line at 0.50) for all the ML models. The predictive performance of random forest model was higher than other ML models, though not significantly (p > 0.01). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel performances were not significantly different from one another. The boxplot shows quartiles at the box ends and the median as the horizontal line in the box. The whiskers show the farthest points that were not outliers. Outliers were defined as those data points that are not within 1.5 times the interquartile ranges. 
\newpage
\includegraphics[height=17cm, width=12cm]{Figure_3.png}

**Figure 3. Interpretation of the linear ML models.** The ranks of absolute feature weights of (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression, were ranked from highest rank, 1, to lowest rank, 100, for each data-split. The feature ranks of the 20 highest ranked OTUs based on their median ranks (median shown in black) are reported here.  OTUs that were associated with classifying a subject as being healthy had negative signs and were shown in blue. OTUs that were associated with classifying a subject having an SRN had positive signs and were shown in red. 
\newpage
\includegraphics{Figure_4.png}

**Figure 4. Interpretation of the non-linear ML models.** (A) SVM with radial basis kernel, (B) decision tree, (C) random forest, and (D) XGBoost feature importances were explained using permutation importance on the held-out test data set. The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation. The 20 OTUs that caused the largest decrease in the AUROC when permuted are reported here. The colors of the box plots represent the OTUs that were shared among the different models; yellow were OTUs that were shared among all the non-linear models, green were OTUs that were shared among the tree-based models, green were the OTUs shared among SVM with radial basis kernel, decision tree and XGBoost, pink were the OTUs shared among SVM with radial basis kernel and XGBoost only, red were the OTUs shared among random forest and XGBoost only and blue were the OTUs shared among decision tree and random forest only. For all of the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance.
\newpage
\includegraphics{Figure_5.png}

@@

\includegraphics{Figure_S1.png}
**Figure S1. Hyperparameter setting performances for linear models.** (A) L2-regularized logistic regression, (B) L1-regularized SVM with linear kernel, and (C) L2-regularized SVM with linear kernel mean cross-validation AUROC values when different hyperparameters were used in training the model. The stars represent the highest performing hyperparameter setting for each model. 
\newpage
\includegraphics[height=30cm, width=15cm]{Figure_S2.png}

**Figure S2. Hyperparameter setting performances for non-linear models.** (A) Decision tree, (B) random forest, (C) SVM with radial basis kernel, and (D) XGBoost mean cross-validation AUROC values when different hyperparameters were used in training the model.  The stars represent the highest performing hyperparameter setting for the models. 
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S3.png}

**Figure S3. Histogram of AUROC differences between L2-regularized logistic regression and random forest for each of the hundred data-splits.**  In 75% of data-splits, the AUROC of random forest was greater than that of L2-regularized logistic regression. The p-value was manually calculated using the sampling distribution of the test statistic under the null hypothesis. We tested how often random forest performed more accurately than L2-regularized logistic regression. The null hypothesis is that the distribution of the difference between the AUROC values of random forest and L2 logistic regression is symmetric about 0, therefore the p-value was calculated for a double-tail event.
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}

**Figure S4. Classification performance of ML models across cross-validation when trained on a subset of the dataset.** (A) L2-regularized logistic regression and (B) random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects. The range among the cross-validation AUROC values within both models at smaller sample sizes were much larger than when the full collection of samples was used to train and validate the models but included the ranges observed with the more complete datasets.
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S5.png}

**Figure S5. NMDS ordination of Bray-Curtis distances.** NMDS ordination relating the community structures of the fecal microbiota from 490 patients (261 patients with normal colonoscopies and 229 patients who have screen relevant neoplasias; SRNs).

\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S6.png}

**Figure S6. Permutation importance analysis.** Permutation importance analysis measures the decrease in the predictive performance of the model after we permute (A) a feature’s or (B) a group of correlated features' values, which breaks the relationship between the feature and the diagnosis.

\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S7.png}

**Figure S7. Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.

\newpage
\includegraphics[height=20cm, width=13cm]{Figure_S8.pdf}

**Figure S8. Relative abundances of the 20 most important OTUs in L2-regularized logistic regression and random forest models.** The most important 20 OTUs were chosen for (A) Random forest and (B) L2-regularized logistic regression models by permutation importance and ranking feature coefficients, respectively. The relative abundances of these OTUs were compared based on the diagnosis of the patients. The minimal differences betweeen relative abundances for these OTUs show that it is not possible to differentiate disease vs healthy states by focusing on individual taxa.

>>>>

---
title: '**A framework for effective application of machine learning to a microbiome-based classification problem**'
csl: mbio.csl
fontsize: 11pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
bibliography: references.bib
---

@@

Running title: Machine learning framework applied to a microbiome study

@@

As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose disease. However, the structure of the human microbiome is remarkably variable among individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models. For example it is not possible to classify individuals as having healthy colons or colorectal lesions using Bray-Curtis distances based on the 16S rRNA gene sequences collected from fecal samples [Figure S1]. This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals. Furthermore, a growing number of studies have shown that it is rare for a single bacterial species to be associated with a disease. Instead, subsets of the microbiome account for differences in health. Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population separately [@segata_metagenomic_2011]. Recently, machine learning (ML) models have grown in popularity among microbiome researchers as our ability to sample large numbers of individuals has grown; such models can effectively account for the interpersonal microbiome variation and the ecology of disease.

@@

In this study, we provided steps toward standardization of machine learning methods for microbiome studies which are often poorly documented and executed. To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing the predictive performance, interpretability and training times of seven modeling approaches with the same dataset and pipeline. We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel. We also trained four non-linear models: SVM with radial basis function kernel, a decision tree, random forest, and gradient boosted trees. We compared their predictive performance, interpretability, and training time. To demonstrate the performance of these modeling approaches and our pipeline on a case study, we used a data from a previously published study that sought to classify individuals as having healthy colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016]. This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. With this dataset, we developed an ML pipeline that can be used in many different scenarios for training and evaluating models. This framework can be easily applied to other host-associated and environmental microbiome datasets. We also provided an aspirational rubric for evaluating the rigor of ML practices applied to microbiome data [Table S1] to urge the audience to be diligent with with their study design, model selection, development, evaluation and interpretation.

@@

Model selection requires tuning hyperparameters. Hyperparameters are parameters that need to be specified or tuned by the user, in order to train a model for a specific modeling problem. For example, when using regularization, C is a hyperparameter that indicates the penalty for overfitting. Hyperparameters are tuned using the training data to find the best model. We selected hyperparameters by performing repeated five-fold cross-validation (CV) on the training set [Figure 1]. The five-fold CV was also stratified to maintain the overall case and control distribution. We chose the hyperparameter values that led to the best average CV predictive performance using the area under the receiver operating characteristic curve (AUROC) [Figure S2 and S3]. The AUROC ranges from 0, where the model’s predictions are perfectly incorrect, to 1.0, where the model perfectly distinguishes between cases and controls. An AUROC value of 0.5 indicates that the model's predictions are no different than random. To select hyperparameters, we performed a grid search for hyperparameter settings when training the models. Default hyperparameter settings in developed ML packages available in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and need to be optimized for each new ML task. For example, L1-regularized SVM with linear kernel showed large variability between different regularization strengths (C) and benefited from tuning as the default C parameter was 1 [Figure S2].

@@

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having healthy colons or SRNs [Figure 2]. The predictive performance of random forest model was higher than other ML models with a median  `r paste(median_iqr(all, "Random_Forest")[1])` [IQR `r paste(median_iqr(all, "Random_Forest")[2])`-`r paste(median_iqr(all, "Random_Forest")[3])`], though not significantly (p=0.5; The p-value was manually calculated using the sampling distribution of the test statistic under the null hypothesis) (Figure S4). Similarly, L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(median_iqr(all, "L2_Logistic_Regression")[1])` [IQR `r paste(median_iqr(all, "L2_Logistic_Regression")[2])`-`r paste(median_iqr(all, "L2_Logistic_Regression")[3])`], `r paste(median_iqr(all, "XGBoost")[1])` [IQR `r paste(median_iqr(all, "XGBoost")[2])`-`r paste(median_iqr(all, "XGBoost")[3])`], `r paste(median_iqr(all, "L2_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L2_Linear_SVM")[2])`-`r paste(median_iqr(all, "L2_Linear_SVM")[3])`] and `r paste(median_iqr(all, "RBF_SVM")[1])` [IQR `r paste(median_iqr(all, "RBF_SVM")[2])`-`r paste(median_iqr(all, "RBF_SVM")[3])`], respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(median_iqr(all, "L1_Linear_SVM")[1])` [IQR `r paste(median_iqr(all, "L1_Linear_SVM")[2])`-`r paste(median_iqr(all, "L1_Linear_SVM")[3])`] and `r paste(median_iqr(all, "Decision_Tree")[1])` [IQR `r paste(median_iqr(all, "Decision_Tree")[2])`-`r paste(median_iqr(all, "Decision_Tree")[3])`], respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and L2-regularized SVM with linear kernel) performed nearly as well as non-linear models.

@@

To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S5]. The variation in cross-validation performance within both models at smaller sample sizes was larger than when the full collection of samples was used to train and validate the models. Because of the high dimensionality of the microbiome data (6920 OTUs), large sample sizes can lead to better models.

__Interpretation of each ML model.__  We often use ML models not just to predict a health outcome, but also to identify potential biomarkers for disease. Therefore, model interpretation becomes crucial for microbiome studies. Interpretability is related to the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017; @ribeiro_why_2016; @nori_interpretml:_2019]. ML models often decrease in interpretability as they increase in complexity. In this study, we used two methods to help interpret our models.

@@

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S7]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S7], 17 of the 20 top OTUs (e.g., OTU 609, 822, 1239) were deemed important by both interpretation methods. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods. These results indicate that both methods are consistent in selecting the most important OTUs.

@@

**Statistical analysis workflow.** Data summaries, statistical analysis, and data visualizations were performed using R (v.3.5.0) with the tidyverse package (v.1.2.1). We compared the performance of the models pairwise by calculating the difference between AUROC values from the same data-split (for 100 data-splits). We determined if the models were significantly different by calculating the empirical p-value  (2 x min(% of AUROC differences $\geq$ 0, % of AUROC differences $\leq$ 0) for the double tail event (e.g., Figure S4).

@@

\newpage
\includegraphics{Figure_1.pdf}
**Figure 1. Machine learning pipeline.**   We split the data to create a training (80%) and held-out test set (20%). The splits were stratified to maintain the overall class distribution. We performed five-fold cross-validation on the training data to select the best hyperparameter setting and then used these hyperparameters to train the models. The model was evaluated on the held-out data set. Abbreviations: cvAUC, cross-validation area under the receiver operating characteristic curve.
\newpage
\includegraphics{Figure_2.tiff}

**Figure 2. Generalization and classification performance of ML models using AUROC values of all cross-validation and testing performances.** The median AUROC for diagnosing individuals with SRN using bacterial abundances was higher than chance (depicted by a horizontal line at 0.50) for all the ML models. The predictive performance of random forest model was higher than other ML models, though not significantly (p > 0.01). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel performances were not significantly different from one another. The boxplot shows quartiles at the box ends and the median as the horizontal line in the box. The whiskers show the farthest points that were not outliers. Outliers were defined as those data points that are not within 1.5 times the interquartile ranges.
\newpage
\includegraphics[height=17cm, width=12cm]{Figure_3.tiff}

**Figure 3. Interpretation of the linear ML models.** The ranks of absolute feature weights of (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression, were ranked from highest rank, 1, to lowest rank, 100, for each data-split. The feature ranks of the 20 highest ranked OTUs based on their median ranks (median shown in black) are reported here.  OTUs that were associated with classifying a subject as being healthy had negative signs and were shown in blue. OTUs that were associated with classifying a subject having an SRN had positive signs and were shown in red.
\newpage
\includegraphics{Figure_4.tiff}

**Figure 4. Interpretation of the non-linear ML models.** (A) SVM with radial basis kernel, (B) decision tree, (C) random forest, and (D) XGBoost feature importances were explained using permutation importance on the held-out test data set. The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation. The 20 OTUs that caused the largest decrease in the AUROC when permuted are reported here. The colors of the box plots represent the OTUs that were shared among the different models; yellow were OTUs that were shared among all the non-linear models, green were OTUs that were shared among the tree-based models, green were the OTUs shared among SVM with radial basis kernel, decision tree and XGBoost, pink were the OTUs shared among SVM with radial basis kernel and XGBoost only, red were the OTUs shared among random forest and XGBoost only and blue were the OTUs shared among decision tree and random forest only. For all of the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance.
\newpage
\includegraphics{Figure_5.tiff}

@@

\end{tabular}

\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S1.tiff}

**Figure S1. NMDS ordination of Bray-Curtis distances.** NMDS ordination relating the community structures of the fecal microbiota from 490 patients (261 patients with normal colonoscopies and 229 patients who have screen relevant neoplasias; SRNs).

\newpage
\includegraphics{Figure_S2.tiff}
**Figure S2. Hyperparameter setting performances for linear models.** (A) L2-regularized logistic regression, (B) L1-regularized SVM with linear kernel, and (C) L2-regularized SVM with linear kernel mean cross-validation AUROC values when different hyperparameters were used in training the model. The stars represent the highest performing hyperparameter setting for each model.

\newpage
\includegraphics[height=30cm, width=15cm]{Figure_S3.tiff}
**Figure S3. Hyperparameter setting performances for non-linear models.** (A) Decision tree, (B) random forest, (C) SVM with radial basis kernel, and (D) XGBoost mean cross-validation AUROC values when different hyperparameters were used in training the model.  The stars represent the highest performing hyperparameter setting for the models.

\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S4.tiff}
**Figure S4. Histogram of AUROC differences between L2-regularized logistic regression and random forest for each of the hundred data-splits.**  In 75% of data-splits, the AUROC of random forest was greater than that of L2-regularized logistic regression. The p-value was manually calculated using the sampling distribution of the test statistic under the null hypothesis. We tested how often random forest performed more accurately than L2-regularized logistic regression. The null hypothesis is that the distribution of the difference between the AUROC values of random forest and L2 logistic regression is symmetric about 0, therefore the p-value was calculated for a double-tail event.

\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S5.tiff}
**Figure S5. Classification performance of ML models across cross-validation when trained on a subset of the dataset.** (A) L2-regularized logistic regression and (B) random forest models were trained using the original study design with 490 subjects and subsets of it with 15, 30, 60, 120, and 245 subjects. The range among the cross-validation AUROC values within both models at smaller sample sizes were much larger than when the full collection of samples was used to train and validate the models but included the ranges observed with the more complete datasets.

\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S6.tiff}
**Figure S6. Permutation importance analysis.** Permutation importance analysis measures the decrease in the predictive performance of the model after we permute (A) a feature’s or (B) a group of correlated features' values, which breaks the relationship between the feature and the diagnosis.

\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S7.tiff}
**Figure S7. Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.

\newpage
\includegraphics[height=20cm, width=13cm]{Figure_S8.tiff}
**Figure S8. Relative abundances of the 20 most important OTUs in L2-regularized logistic regression and random forest models.** The most important 20 OTUs were chosen for (A) Random forest and (B) L2-regularized logistic regression models by permutation importance and ranking feature coefficients, respectively. The relative abundances of these OTUs were compared based on the diagnosis of the patients. The minimal differences betweeen relative abundances for these OTUs show that it is not possible to differentiate disease vs healthy states by focusing on individual taxa.
